\textbf{Linear Algebra -- Part II}\\A summary for MIT 18.06SC

### 2020-07-16 skills/math/linear_algebra/linear_algebra_summary_p2.tex
update

\subsection{Orthogonal vectors and subspaces}
\begin{defn}[\textit{Orthogonal vectors}]\label{def:vortho}
	Consider two vectors $x,y \in \bbR^{n}$. The vectors $x$ and $y$ are orthogonal,  denoted $x \perp y$, if and only if $x^T y = y^T x = 0$.
\end{defn}

\begin{defn}[\textit{Orthogonal subspaces}]\label{def:sortho}
	Consider two subspaces $S,T$. The subspaces $S$ and $T$ are orthogonal,  denoted $S \perp T$, if and only if for any $s \in S$ and $ t \in T$, $s^T t = t^T s = 0$.
\end{defn}

Given a matrix $\mA \in \bbR^{m \times n}$, there are four subspaces related to $\mA$\colon column space $C(\mA)$, row space $C(\mA^T)$, nullspace $N(\mA)$, and left nullspace $N(\mA^T)$. Suppose the rank of $\mA$ is $rank(\mA) = r$, the dimensions of these subspaces are: 
\[ dim(C(\mA)) = dim(C(\mA^T)) = r,\quad  dim(N(A)) = n-r,\quad dim(N(\mA^T)) = m-r. \]

\begin{thm}[Orthogonality of matrix subspaces]\label{thm:ortho}
	Consider a matrix $\mA \in \bbR^{m \times n}$. The row space $C(\mA^T)$ and the nullspace $N(\mA)$ are orthogonal; the column space $C(\mA)$ and the left nullspace $N(\mA^T)$ are orthogonal, i.e.\
	\[ C(\mA^T) \perp N(\mA); \quad C(\mA) \perp N(\mA^T).  \]
\end{thm}

\begin{proof}
	Consider the matrix $\mA \in \bbR^{m \times n}$. For any vector $x \in N(\mA)$, we have $\mA x = 0$. 
	\begin{align}
		\mA x  = \begin{bmatrix}
			a_1^T x \\ \vdots \\ a_m^T x 
		\end{bmatrix} = \begin{bmatrix}
			0\\ \vdots \\0
		\end{bmatrix},
	\end{align}
	where $a_i \in \bbR^n, i \in [m]$ is the $i$-th  row of $\mA$. By definition~\ref{def:vortho}, $x$ is orthogonal to the rows in matrix $\mA$. For any vector $v \in C(\mA^T)$, $v$ is a linear combination of the rows $v = c_1 a_1 + ...+c_m a_m$, where $c_i, i\in [m]$ are constants. Multiplying vectors $v$ and $x$, we have
	\[  v^T x =  c_1 a_1^T x + \cdots + c_m a_m^T x = 0.  \]
	Therefore, $v \perp x$, and $N(\mA) \perp C(\mA^T)$.

@@

\begin{thm}[Relationship between $\mA^T \mA$ and $\mA$]\label{thm:aa}
	Consider a matrix $\mA \in \bbR^{m \times n}$. We have
	\begin{equation}
		N(\mA^T \mA) = N(\mA) ; \quad rank(\mA^T \mA) = rank(\mA).
	\end{equation}
\end{thm}
\begin{proof}
	Consider a matrix $\mA \in \bbR^{m \times n}$. First, we prove that $N(\mA^T \mA) = N(\mA)$.

	On one hand, if $x \in N(\mA)$, then $\mA x = 0 \Rightarrow \mA^T  \mA x = \mA^T 0 = 0$. Therefore, for any $x \in N(\mA)$, the vector $x \in N(\mA^T \mA)$.

	On the other hand, we  prove that for any $x \in N(\mA^T \mA)$ the vector $x \in N(\mA)$ by contradiction.

	Suppose there is a vector $x \in N(\mA^T \mA^T)$ but $x \notin N(\mA)$. We have
	\begin{equation}\label{eq:aaa}
		\mA x = b \neq 0, \quad \mA^T \mA x = 0 \quad \Rightarrow \quad \mA^T b = 0.
	\end{equation} 
	By the first equality in \eqref{eq:aaa}, $b \in C(\mA)$, and by the third equation in \eqref{eq:aaa}, $ b \in N(\mA^T)$. This contradicts the Theorem~\ref{thm:ortho}, $C(\mA) \perp N(\mA^T)$. Therefore, for any $x \in N(\mA^T \mA)$, the vector $x \in N(\mA)$.

@@

\begin{cor}[Invertibility of $\mA^T \mA$]\label{cor:invert}
	If $\mA$ has independent columns, then $\mA^T \mA$ is invertible.
\end{cor}
\begin{proof}
	If $\mA \in \bbR^{m \times n}$ has independent columns, then $rank(\mA) = n$. By Theorem~\ref{thm:aa}, $rank(\mA^T \mA) = rank(\mA) = n$. Since $\mA^T \mA \in \bbR^{n \times n}$ is a square matrix, $\mA^T \mA$ is invertible.
\end{proof}

\subsection{Projections onto subspaces}
\begin{defn}[\textit{Projection and projection matrix}]\label{def:project}
	Given a vector $x \in \bbR^{n}$ and a matrix $\mA^{m \times n}$ with independent columns, the vector $p \in C(\mA)$, such that
	\begin{equation}\label{eq:project}
		(x-p) \perp C(\mA),
	\end{equation}

	is the projection of vector $x$ onto the space $C(\mA)$. Since $p \in C(\mA)$, there exists a vector $\hat x$ such that $p = \mA \hat x$. By equation~\eqref{eq:project}, we have
	\[ \mA^T (x - p) = \mA^T(x - \mA \hat x) = 0  \quad  \Rightarrow\quad  \hat x = (\mA^T \mA)^{-1} \mA^T x,\quad p = \mA(\mA^T \mA)^{-1} \mA^T x.  \] 
	The matrix $\mP \overset{\Delta}{=}   \mA(\mA^T \mA)^{-1} \mA^T$ is called a projection matrix.
\end{defn}

\begin{thm}[Properties of projection matrix]\label{thm:projectm}
	Consider a projection matrix $\mP$ to the column space $C(\mA)$, where $\mA \in \bbR^{m \times n}$ is a matrix. Then,
	\begin{equation}\label{eq:projectm}
		\mP^T = \mP;\quad  \mP^2 = \mP.
	\end{equation}
\end{thm}
\begin{proof}
	Since the projection matrix $\mP =   \mA(\mA^T \mA)^{-1} \mA^T$ and $\mA^T \mA$ is symmetric, then
	\[  \mP^T = (\mA(\mA^T \mA)^{-1} \mA^T )^T = \mA (\mA^T \mA)^{-1} \mA^T = \mP. \]
	\[  \mP^2 = \mP^T \mP =  \mA (\mA^T \mA)^{-1} \mA^T  \mA (\mA^T \mA)^{-1} \mA^T =  \mA (\mA^T \mA)^{-1} \mA^T = \mP. \]
\end{proof}

\begin{cor}[Projection onto $N(\mA^T)$]\label{cor:iprojectm}
	Suppose $\mP$ is a projection matrix in Theorem~\ref{thm:projectm}, then $\mI-\mP$ is also a projection matrix to the left nullspace $N(\mA^T)$. 
\end{cor}

\begin{proof}
	For any $x \in \bbR^n$, we have
	\[ x - \mP x \perp C(\mA) \quad \Rightarrow \quad (I-\mP)x \perp C(\mA) \quad \Rightarrow \quad  (\mI-\mP)x \in N(\mA^T) \text{ and } (x - (\mI-\mP)x) \perp N(\mA^T). \]
	Therefore, $\mI-\mP$ is a projection matrix to the left nullspace $N(\mA^T)$.
\end{proof}

\subsection{Projection matrices and least squares}
Given an observation vector $y \in \bbR^n$ and a design matrix $\mX \in \bbR^{n \times (k+1)}$, we propose the linear regression model
\[ y = \mX \beta +\epsilon,\]
where $\beta = (\beta_0,\beta_1,...,\beta_k)$ are regression coefficients and $\epsilon$ is the noise. The least square estimate of $\beta$ is the minimizer of the loss, i.e.\
\[  \hat \beta_{LS} = \argmin_{\beta \in \bbR^{k+1}} \onorm{y - \mX \beta}^2, \]
where $\onorm{\cdot}$ is the euclidean norm. The vector $\mX 
\hat \beta_{LS}$ can be considered as a the projection of $y$ onto the column space of $\mX$, which minimizes the distance from $y$ to the column space $C(\mX)$. Therefore, we may use projection tools to solve the least square estimate.

The projection $\mX \hat \beta_{LS}$ satisfies
\[ \mX^T( y - \mX \hat \beta_{LS} ) = 0 \quad \Rightarrow \quad \hat \beta_{LS} = (\mX^T \mX)^{-1} \mX^T y. \]

The estimate $\hat \beta_{LS}$ consists with the estimates solved by using the derivative.

@@

\begin{defn}[\textit{Orthonormal matrix and orthogonal matrix}]\label{def:orthonom}
	Consider a matrix $\mQ \in \bbR^{m \times n}$. If the columns of $\mQ$ are orthonormal, the matrix $\mQ$ is an orthonormal matrix. If $ m= n$, the square matrix $\mQ$ is an orthogonal matrix.
\end{defn}

If $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix, $\mQ^T \mQ = \mI_n$.  If $\mQ$ is an orthogonal matrix, $\mQ^T = \mQ^{-1}$.
For orthonormal matrix $\mQ$, the projection matrix to $C(\mP)$  becomes $\mP = \mI_m$.

\begin{defn}[\textit{Gram-Schimidt Process and QR decomposition}]\label{gram-schimidt}
Consider a matrix $\mA \in \bbR^{m \times n}$ with $rank(\mA) = n$. Gram-Schimidt process finds the orthonormal basis for $C(\mA)$. Let $a_i \in \bbR^{m}, i \in [m]$ be the columns of matrix $\mA$. 
\begin{align}
	&u_1 = a_1, &&\quad e_1 = \frac{u_1}{\onorm{u_1}}\\
	&u_2 = a_2 - \frac{u_1^T a_2}{u_1^T u_1} u_1 ,&&\quad  e_2 = \frac{u_2}{\onorm{u_2}}\\
	&u_3 = a_3 - \frac{u_1^T a_3}{u_1^T u_1} u_1 - \frac{u_2^T a_3}{u_2^T u_2} u_2,&&\quad  e_2 = \frac{u_3}{\onorm{u_3}}\\
	\vdots
\end{align}
 The vectors $e_1,...,e_n$ are orthonormal basis of the $C(\mA)$. By matrix operations, we obtain a decomposition of matrix $\mA$
 \begin{equation}\label{eq:qr}
 	\mA = [a_1,...,a_n] = [e_1,....,e_2] \begin{bmatrix}
 		e_1^T a_1 & e_1^T a_2 & \cdots & e_1^T a_n\\
 		0&e_2^T a_2 & \cdots & e_2^T a_n\\
 		\vdots &\vdots & & \vdots \\
 		0 &0&\cdots & e_n^T a_n
 	\end{bmatrix} = \mQ \mR
 \end{equation}
	where $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix and $\mR \in \bbR^{n \times n}$ is an upper triangular matrix. We call the matrix decomposition in equation~\eqref{eq:qr} QR decomposition.
\end{defn}

@@

\subsection{Properties of determinants}
We give ten properties of determinants. The last seven properties are deduced by first three basic properties.
\begin{enumerate}
	\item\label{prop1} The determinant of identity matrix is equal to 1, i.e.\ $\det(\mI) = 1$.
	\item\label{prop2} Exchanging two rows of a matrix reverses the sign of the matrix's determinant. Hence, an odd number of row exchanges reverse the sign of the determinant while an even number of row exchanges maintain the sign of the determinant.
	\item\label{prop3} \begin{enumerate} 
	\item[(a)] If one row of the matrix is multiplied by a constant $t$, the determinant of the matrix is multiplied by $t$.
	\begin{align}
		\begin{vmatrix}
			ta&tb\\c &d 
		\end{vmatrix} = t \begin{vmatrix}
			a&b\\c &d
			\end{vmatrix}
	\end{align} 
	\item[(b)] The determinant is linearly additive on the rows of the matrix.
	\begin{align}
		\begin{vmatrix}
			a+a'&b+b'\\c &d 
		\end{vmatrix} =  \begin{vmatrix}
			a&b\\c &d
			\end{vmatrix} +  \begin{vmatrix}
			a'&b'\\c &d
			\end{vmatrix} 
	\end{align}
	\end{enumerate}

	\item \label{propidenticalrow} If two rows of a matrix are equal, the determinant of the matrix is 0.  
	\begin{proof}
		Suppose $\mA$ has two identical rows $a_i, a_j$. The matrix after exchanging $a_i, a_j$, $\mA'$, is the same as the original matrix $\mA$. By property~\ref{prop2}, $\det(\mA) = -\det(\mA') = -\det(\mA)$. Therefore, $\det(\mA) = 0$.
	\end{proof}

	\item \label{propelim} For a square matrix $\mA \in \bbR^{n \times n}$, subtracting $t$ times of the $i$-th row from the $j$-th row does not change the determinant of $\mA$, where $i \neq j$.
	\begin{proof}
		By property~\ref{prop3} and property~\ref{propidenticalrow}, 
	\begin{align}
		 \begin{vmatrix}
			a & b \\ c - ta&d -tb
		\end{vmatrix} = \begin{vmatrix}
			a & b \\ c &d
		\end{vmatrix} -t \begin{vmatrix}
			a & b \\ a& b
		\end{vmatrix}  = \begin{vmatrix}
			a & b \\ c &d.
		\end{vmatrix}
	\end{align}
	The proof for higher-dimension matrix is similar.
	\end{proof}

	\item\label{prop0row} If the matrix $\mA$ has a row that is all zeros, the determinant of $\mA$ is 0.
	\begin{proof}
		By property~\ref{prop3}, letting $t = 0$ leads to property~\ref{prop0row}.
	\end{proof}

	\item\label{propdiag} Suppose $\mA \in \bbR^{n \times n}$ is a triangular matrix with diagonal elements $d_1,...,d_n$. Then, the determinant $\det(\mA) = \prod_{i = 1}^n d_i$. 
	\begin{proof}
		By property~\ref{propelim}, eliminating the triangular matrix $\mA$ to a diagonal matrix does not change the determinant of $\mA$. By property~\ref{prop3} and \ref{prop1}, $\det(\mA) = d_1 ...d_n \det(\mI_n) = d_1 ...d_n $. 
	\end{proof}

	\item\label{propsingular} If the square matrix $\mA$ is singular, the determinant of $\mA$ is 0.

	\begin{proof}
		If $\mA$ is a singular matrix, the reduced row echelon form of $\mA$, RREF($\mA$), has a row with all 0 entries. By property~\ref{propelim} and \ref{prop0row}, the determinant of $\mA$ is 0.
	\end{proof}

	\item\label{propproduct} For square matrices $\mA, \mB \in \bbR^{n \times n}$, the determinant of the product satisfies $\det(\mA \mB) = \det(\mA) \det(\mB).$

	\begin{proof}
		If at least one of $\mA, \mB$ is singular, the product $\mA \mB$ is also singular. Then, $\det(\mA \mB) = \det(\mA) \det(\mB) = 0$. If both $\mA, \mB$ are invertible, $\mA, \mB$ can be eliminated to diagonal matrices $\mD_{A}, \mD_{B}$ by elimination. Suppose $a_1,...,a_n$ and $b_1,...,b_n$ are diagonal elements for $\mD_A$ and $\mD_B$ respectively. By property~\ref{propdiag}, $\det(AB) = \prod_{i=1}^n a_i b_i = \det(\mA) \det(\mB)$. 
	\end{proof}

	\item\label{proptrans} For a square matrix $\mA$, the determinant $\det(\mA^T) = \det(\mA)$.

	\begin{proof}
		By LU decomposition, the matrix $\mA = \mL \mU$, where $\mL$ is a lower-triangular matrix and $\mU$ is an upper-triangular matrix. The transport $\mL^T ,\mU^T$ have the same diagonal elements with $\mL, \mU$ respectively. By property~\ref{propdiag}, $\det(\mL^T)= \det(\mL),  \det(\mU^T) = \det(\mU)$. Therefore, by property~\ref{propproduct}, $\det(\mA) = \det(\mL) \det(\mU) = \det(\mU^T) \det(\mL^T) = \det(\mU^T \mL^T) = \det(\mA^T)$. 
			\end{proof}
\end{enumerate}

\subsection{Determinant computation}
\begin{prop}[Big formula for computing determinant]
	Consider a square matrix $\mA = \entry{a_{ij}} \in \bbR^{n \times n}$. The big formula for computing the determinant of $\mA$ is
	\[ \det(\mA) =  \sum_{(\alpha_1,...,\alpha_n) \in \tP} (-1)^{N(\alpha_1,...,\alpha_n)} a_{1,\alpha_1} ... a_{n \alpha_n},  \]
	where $(\alpha_1, \alpha_2, ...,\alpha_n)$ is the permutation of $(1,2,...,n)$, $\tP$ is the collection of all possible  $(\alpha_1,...,\alpha_n)$, and $N(\alpha_1,...,\alpha_n)$ is the number of necessary exchanges from $(\alpha_1,\alpha_2,...,\alpha_n)$  to $(1,2,...,n)$. 
\end{prop}

>>>>

\subsection{Orthogonal vectors and subspaces}
\begin{defn}[\textit{Orthogonal vectors}]\label{def:vortho}
	Let  $x,y \in \bbR^{n}$ be two vectors. The vectors $x$ and $y$ are orthogonal,  denoted $x \perp y$, if and only if $x^T y = y^T x = 0$.
\end{defn}

\begin{defn}[\textit{Orthogonal subspaces}]\label{def:sortho}
	Let  $S,T$ be two subspaces $S,T$. The subspaces $S$ and $T$ are orthogonal,  denoted $S \perp T$, if and only if for any $s \in S$ and $ t \in T$, $s^T t = t^T s = 0$.
\end{defn}

Let $\mA \in \bbR^{m \times n}$ be a matrix. There are four subspaces related to $\mA\colon$ column space $C(\mA)$, row space $C(\mA^T)$, nullspace $N(\mA)$, and left nullspace $N(\mA^T)$. Suppose the rank of $\mA$ is $rank(\mA) = r$, the dimensions of these subspaces are: 
\[ dim(C(\mA)) = dim(C(\mA^T)) = r,\quad  dim(N(A)) = n-r,\quad dim(N(\mA^T)) = m-r. \]

\begin{thm}[Orthogonality of matrix subspaces]\label{thm:ortho}
	Let  $\mA \in \bbR^{m \times n}$ be a matrix. The row space $C(\mA^T)$ and the nullspace $N(\mA)$ are orthogonal. The column space $C(\mA)$ and the left nullspace $N(\mA^T)$ are orthogonal; i.e.,
	\[ C(\mA^T) \perp N(\mA)\quad  \text{and} \quad C(\mA) \perp N(\mA^T).  \]
\end{thm}

\begin{proof}
	For any vector $x \in N(\mA)$, we have $\mA x = 0$. Specifically,
	\begin{align}
		\mA x  = \begin{bmatrix}
			a_1^T x \\ \vdots \\ a_m^T x 
		\end{bmatrix} = \begin{bmatrix}
			0\\ \vdots \\0
		\end{bmatrix},
	\end{align}
	where $a_i \in \bbR^n$ is the $i$-th  row of $\mA$, for all $ i \in [m]$. By Definition~\ref{def:vortho}, $x$ is orthogonal to the rows in matrix $\mA$. For any vector $v \in C(\mA^T)$, $v$ is a linear combination of the rows $v = c_1 a_1 + ...+c_m a_m$, where $c_i \in \bbR$, for all $i\in [m]$. Taking inner product between vectors $v$ and $x$, we have
	\[  v^T x =  c_1 a_1^T x + \cdots + c_m a_m^T x = 0.  \]
	Therefore, $v \perp x$, and $N(\mA) \perp C(\mA^T)$.

@@

\begin{thm}[Relationship between $\mA^T \mA$ and $\mA$]\label{thm:aa}Let $\mA \in \bbR^{m \times n}$ be a matrix. We have
	\begin{equation}
		N(\mA^T \mA) = N(\mA) ; \quad rank(\mA^T \mA) = rank(\mA).
	\end{equation}
\end{thm}
\begin{proof}
	Fxirst, we prove that $N(\mA^T \mA) = N(\mA)$.

	On one hand, if $x \in N(\mA)$, then $\mA x = 0$, which implies that $ \mA^T  \mA x = \mA^T 0 = 0$. Therefore, for any $x \in N(\mA)$, the vector $x \in N(\mA^T \mA)$.

	On the other hand, we  prove  by contradiction that for any $x \in N(\mA^T \mA)$ the vector $x \in N(\mA)$.

	Suppose there is a vector $x \in N(\mA^T \mA^T)$ but $x \notin N(\mA)$. We have
	\begin{equation}\label{eq:aaa}
		\mA x = b \neq 0, \quad \mA^T \mA x = 0 \quad \Rightarrow \quad \mA^T b = 0.
	\end{equation} 
	By the first equality in \eqref{eq:aaa}, $b \in C(\mA)$, and by the third equation in \eqref{eq:aaa}, $ b \in N(\mA^T)$. This contradicts the Theorem~\ref{thm:ortho} that $C(\mA) \perp N(\mA^T)$. Therefore, for any $x \in N(\mA^T \mA)$, the vector $x \in N(\mA)$.

@@

\begin{cor}[Invertibility of $\mA^T \mA$]\label{cor:invert}
	If $\mA$ has independent columns, then $\mA^T \mA$ is invertible.
\end{cor}
\begin{proof}
Suppose $\mA \in \bbR^{m \times n}$ is a matrix with independent columns; i.e., $rank(\mA) = n$. By Theorem~\ref{thm:aa}, $rank(\mA^T \mA) = rank(\mA) = n$. Since $\mA^T \mA \in \bbR^{n \times n}$ is a square matrix, $\mA^T \mA$ is invertible.
\end{proof}

\subsection{Projections onto subspaces}
\begin{defn}[\textit{Projection and projection matrix}]\label{def:project} Let  $x \in \bbR^{n}$ be a vector and $\mA^{n \times m}$ be a matrix with independent columns. The vector $p \in C(\mA)$ that satisfies
	\begin{equation}\label{eq:project}
		(x-p) \perp C(\mA),
	\end{equation} 
	is called the projection of vector $x$ onto the column space $C(\mA)$. For all $x \in \bbR^{n}$ and the corresponding projection $p$, the matrix $\mP \in \bbR^{m \times m}$ that satisfies
	\begin{equation}
		p = \mP x,
	\end{equation}
	is called the projection matrix of $\mA$ from $\bbR^n$ onto the column space  $C(\mA)$.
\end{defn}

\begin{prop}[Projection matrix of $C(\mA)$]\label{prop:project}
	Let  $x \in \bbR^{n}$ be a vector and $\mA^{m \times n}$ be a matrix with independent columns. The projection matrix of $\mA$ from $\bbR^n$ onto the column space  $C(\mA)$ is 
	\begin{equation}
		\mP =  \mA (\mA^T \mA)^{-1} \mA^T.
	\end{equation}
\end{prop}

\begin{proof}
	By Definition~\ref{def:project}, the projection $p \in C(\mA)$. Then, there exists a  vector $\hat x \in \bbR^m$ such that $p = \mA \hat x$.  By equation~\eqref{eq:project}, for all $x \in \bbR^n$, we have 
\[ A^T (x - \mA \hat x) = 0 \quad \Rightarrow  \quad \hat x = (\mA^T \mA)^{-1} \mA^T x \quad \Rightarrow \quad p = \mA (\mA^T \mA)^{-1} \mA^T  x. \]
Therefore, the matrix $\mP =  \mA (\mA^T \mA)^{-1} \mA^T$ is the projection matrix of $\mA$ from $\bbR^n$ onto the column space $C(\mA)$.
\end{proof}

\begin{thm}[Properties of projection matrix]\label{thm:projectm}
Let $\mA \in \bbR^{m \times n}$ be a matrix, and $\mP$ be the projection matrix of $\mA$ onto the column space. Then,
	\begin{equation}\label{eq:projectm}
		\mP^T = \mP;\quad  \mP^2 = \mP.
	\end{equation}
\end{thm}
\begin{proof}
	By Proposition~\ref{prop:project}, the projection matrix is $\mP =   \mA(\mA^T \mA)^{-1} \mA^T$. Then,
	\[  \mP^T = (\mA(\mA^T \mA)^{-1} \mA^T )^T = \mA (\mA^T \mA)^{-1} \mA^T = \mP. \]
	\[  \mP^2 = \mP^T \mP =  \mA (\mA^T \mA)^{-1} \mA^T  \mA (\mA^T \mA)^{-1} \mA^T =  \mA (\mA^T \mA)^{-1} \mA^T = \mP. \]
\end{proof}

\begin{cor}[Projection onto $N(\mA^T)$]\label{cor:iprojectm}
	Suppose $\mP$ is a projection matrix in Theorem~\ref{thm:projectm}. Then $\mI-\mP$ is also a projection matrix of $\mA$ from $\bbR^n$ onto the left nullspace $N(\mA^T)$. 
\end{cor}

\begin{proof}
	By Definition~\ref{def:project}, for any $x \in \bbR^n$, we have
	\[ x - \mP x \perp C(\mA) \quad \Rightarrow \quad (I-\mP)x \perp C(\mA). \]
	By Theorem~\ref{thm:ortho}, the column space of $\mA$ is orthogonal to the left null space of $\mA$. Then,
	\[ (\mI-\mP)x \in N(\mA^T) \quad \text{ and }\quad  (x - (\mI-\mP)x) \perp N(\mA^T). \]
	Therefore, $\mI-\mP$ is a projection matrix of $\mA$ from $\bbR^n$ onto the left nullspace $N(\mA^T)$.
\end{proof}

\subsection{Projection matrices and least squares}
Let $y \in \bbR^n$ be a vector and $\mX \in \bbR^{n \times (k+1)}$ be a design matrix. We propose the linear regression model as
\[ y = \mX \beta +\epsilon,\]
where $\beta = (\beta_0,\beta_1,...,\beta_k)$ are regression coefficients and $\epsilon$ is the noise. The least square estimate of $\beta$ is the minimizer of the loss; i.e.,
\[  \hat \beta_{LS} = \argmin_{\beta \in \bbR^{k+1}} \onorm{y - \mX \beta}^2, \]
where $\onorm{\cdot}$ is the euclidean norm. We consider the vector $\mX 
\hat \beta_{LS}$ as the projection of $y$ onto the column space of $\mX$, which minimizes the distance from $y$ to the column space $C(\mX)$. Therefore, we may use projection to solve the minimization problem.

By Definition~\ref{def:project}, the projection $\mX \hat \beta_{LS}$ satisfies
\[ \mX^T( y - \mX \hat \beta_{LS} ) = 0 \quad \Rightarrow \quad \hat \beta_{LS} = (\mX^T \mX)^{-1} \mX^T y. \]

The estimate $\hat \beta_{LS}$ is consistent with the estimates solved by using the derivative.

@@

\begin{defn}[\textit{Orthonormal matrix and orthogonal matrix}]\label{def:orthonom}
	Let  $\mQ \in \bbR^{m \times n}$ be a matrix. The matrix $\mQ$ is an orthonormal matrix, if the columns of $\mQ$ are orthonormal. When $ m= n$, the square matrix $\mQ$ is an orthogonal matrix.
\end{defn}

Suppose $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix. Then, we have $\mQ^T \mQ = \mI_n$.  When $m = n$, $\mQ$ is an orthogonal matrix and $\mQ^T = \mQ^{-1}$. The projection matrix of $\mQ$ onto the column space $C(\mQ)$, denoted $\mP$,  becomes $\mP = \mI_m$.

\begin{defn}[\textit{Gram-Schimidt Process and QR decomposition}]\label{gram-schimidt}
Let $\mA \in \bbR^{m \times n}$ be a matrix with $rank(\mA) = n$, and  $a_i \in \bbR^{m}$ be the column of matrix $\mA$, for all $i \in [n]$. Gram-Schimidt process finds the orthonormal basis for $C(\mA)$ as following,
\begin{align}
	&u_1 = a_1, &&\quad e_1 = \frac{u_1}{\onorm{u_1}};\\
	&u_2 = a_2 - \frac{u_1^T a_2}{u_1^T u_1} u_1 ,&&\quad  e_2 = \frac{u_2}{\onorm{u_2}};\\
	&u_3 = a_3 - \frac{u_1^T a_3}{u_1^T u_1} u_1 - \frac{u_2^T a_3}{u_2^T u_2} u_2,&&\quad  e_2 = \frac{u_3}{\onorm{u_3}};\\
	\vdots\\
	&u_n = a_n - \sum_{k = 1}^{n-1} \frac{u_k^T a_k}{u_k^T u_k} u_k, && \quad e_n = \frac{u_n}{\onorm{u_n}}.
\end{align}
 The vectors $e_1,...,e_n$ are orthonormal basis of the $C(\mA)$. Based on the basis obtained by Gram-Schimidit, we decompose the matrix $\mA$ as following,
 \begin{equation}\label{eq:qr}
 	\mA = [a_1,...,a_n] = [e_1,....,e_2] \begin{bmatrix}
 		e_1^T a_1 & e_1^T a_2 & \cdots & e_1^T a_n\\
 		0&e_2^T a_2 & \cdots & e_2^T a_n\\
 		\vdots &\vdots & & \vdots \\
 		0 &0&\cdots & e_n^T a_n
 	\end{bmatrix} = \mQ \mR,
 \end{equation}
	where $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix and $\mR \in \bbR^{n \times n}$ is an upper triangular matrix. The matrix decomposition in equation~\eqref{eq:qr} is called QR decomposition.
\end{defn}

@@

\subsection{Properties of determinants}
We give ten properties of determinants. The last seven properties are deduced by the first three basic properties.
\begin{enumerate}
	\item\label{prop1} The determinant of identity matrix is equal to 1; i.e., $\det(\mI) = 1$.
	\item\label{prop2} Exchanging two rows of a matrix reverses the sign of the matrix's determinant. Hence, an odd number of row exchanges reverse the sign of the determinant while an even number of row exchanges maintain the sign of the determinant.
	\item\label{prop3} \begin{enumerate} 
	\item[(a)] If one row of the matrix is multiplied by a constant $t$, the determinant of the matrix is multiplied by $t$.
	\begin{align}
		\begin{vmatrix}
			ta&tb\\c &d 
		\end{vmatrix} = t \begin{vmatrix}
			a&b\\c &d
			\end{vmatrix}
	\end{align} 
	\item[(b)] The determinant is linearly additive on the rows of the matrix.
	\begin{align}
		\begin{vmatrix}
			a+a'&b+b'\\c &d 
		\end{vmatrix} =  \begin{vmatrix}
			a&b\\c &d
			\end{vmatrix} +  \begin{vmatrix}
			a'&b'\\c &d
			\end{vmatrix} 
	\end{align}
	\end{enumerate}

	\item \label{propidenticalrow} If the matrix contains two identical rows, the determinant of the matrix is 0.  
	\begin{proof}
		Suppose $\mA$ has two identical rows $a_i, a_j$. The matrix after exchanging $a_i, a_j$, denoted $\mA'$, is the same as the original matrix $\mA$. By property~\ref{prop2}, we have $\det(\mA) = -\det(\mA') = -\det(\mA)$. Therefore, $\det(\mA) = 0$.
	\end{proof}

	\item \label{propelim} Subtracting $t$ times of the $i$-th row from the $j$-th row does not change the determinant of the matrix, where $i \neq j$.
	\begin{proof}
		Take a two-by-two matrix as an example. By property~\ref{prop3} and property~\ref{propidenticalrow}, 
	\begin{align}
		 \begin{vmatrix}
			a & b \\ c - ta&d -tb
		\end{vmatrix} = \begin{vmatrix}
			a & b \\ c &d
		\end{vmatrix} -t \begin{vmatrix}
			a & b \\ a& b
		\end{vmatrix}  = \begin{vmatrix}
			a & b \\ c &d
		\end{vmatrix}.
	\end{align}
	The proof for higher-dimension matrix is similar.
	\end{proof}

	\item\label{prop0row} If the matrix has a row that is all zeros, the determinant of the matrix is 0.
	\begin{proof}
		By property~\ref{prop3}, letting $t = 0$ leads to property~\ref{prop0row}.
	\end{proof}

	\item\label{propdiag} Let $\mA \in \bbR^{n \times n}$ be a triangular matrix with diagonal elements $d_1,...,d_n$. The determinant $\det(\mA) = \prod_{i = 1}^n d_i$. 
	\begin{proof}
	We eliminate the matrix $\mA$ to a diagonal matrix, denoted $\mA'$. By property~\ref{propelim}, $\det{\mA} = \det{\mA'}$. Since $\mA$ is triangular, the diagonal elements of $\mA'$ are still $d_1,...,d_n$.
 		 By property~\ref{prop3} and \ref{prop1}, $\det(\mA') = \prod_{i = 1}^n d_i\det(\mI_n) =\prod_{i = 1}^n d_i$. Therefore, $\det(\mA) = \prod_{i = 1}^n d_i$.
	\end{proof}

	\item\label{propsingular} If the square matrix is singular, the determinant of the matrix is 0.

	\begin{proof}
		If $\mA$ is a singular matrix, the reduced row echelon form of $\mA$, denoted $RREF(\mA)$, has a row with all 0 entries. By and \ref{prop0row}, $\det(RREF(\mA)) = 0$. Therefore, by property~\ref{propelim}, we have $\det(\mA) =\det(RREF(\mA)) =  0$. 
	\end{proof}

	\item\label{propproduct} Let  $\mA, \mB \in \bbR^{n \times n}$ be two square matrices, the determinant of the matrix product is $\det(\mA \mB) = \det(\mA) \det(\mB).$

	\begin{proof}
		First, if at least one of $\mA, \mB$ is singular, the product $\mA \mB$ is also singular. Then, $\det(\mA \mB) = \det(\mA) \det(\mB) = 0$.

		Second, let $\mE$ be an elimination matrix. The product $\mE \mB$ means implementing a linear operation on the rows of $\mB$. By property~\ref{propelim}, linear operation does not change the determinant of $\mB$. Therefore, we have
		\begin{equation}\label{eq:producteli}
			|\mE \mB| = |\mE| |\mB|
		\end{equation}

		Last, suppose that both $\mA, \mB$ are invertible. By elimination, there exist a sequence of elimination matrix $\{ \mE_i \}_{i = 1}^n$ such that 
		\[ \mE_n \cdots \mE_1 \mA = \mI.  \]
		By Lemma 2 in Linear Algebra-Part I, the inverse of an elimination matrix is still an elimination matrix. Let $\mE'_k$ denote the inverse $\mE^{-1}_k$, for all $k \in [n]$. We rewrite $\mA$ as
		\[ \mA = \mE'_n \cdots \mE'_1. \]
		Combing equation~\eqref{eq:producteli}, we have
		\begin{equation}\label{eq:prod1}
			|\mA \mB| = | \mE'_n \cdots \mE'_1 \mB | = |\mE'_n| |\mE'_{n-1} \cdots \mE'_1 \mB| = \cdots = |\mE'_n|\cdots |\mE'_1||\mB|. 
		\end{equation}
		Applying the equation~\eqref{eq:producteli} again to the term $|\mE'_n|\cdots |\mE'_1| $, we have
		\begin{equation}\label{eq:prod2}
			|\mE'_n|\cdots |\mE'_1|  =|\mE'_n|\cdots |\mE'_3| |\mE'_2 \mE'_1| = \cdots = | \mE'_n \cdots \mE'_1|.
		\end{equation}
		Hence, combing equation~\eqref{eq:prod1} with equation~\eqref{eq:prod2}, we obtain the result 
		\[ |\mA \mB|  = | \mE'_n \cdots \mE'_1| |\mB| = |\mA| |\mB|.\]
	\end{proof}

	\item\label{proptrans} Let $\mA$ be a square matrix, the determinant $\det(\mA^T) = \det(\mA)$.
